---
title: "Generalized additive model of gene expression from U1 snRNP binding-site features"
output:
  html_document:
    toc: true
    number_sections: true
---

This document reproduces the generalized additive model (GAM) analysis used to
relate U1 snRNP binding-site (U1SS) features to mean gene expression across ten
human tissues. Input paths are relative to the project directory so that the
analysis can be shared without user-specific filesystem paths.

Expected input files in `data/`:

- `nearest_10_U1_sites_by_gene.RData`: the output from
  `Find_top-n_nearest_U1sites_publication.R`, containing
  `nearest_U1sitelist`.
- `down1000_sense_score0 or more.txt`: tab-delimited U1SS summary data.
- `rna_tissue_consensus.tsv`: Human Protein Atlas tissue-expression data.

Set the `U1SS_DATA_DIR` or `U1SS_RESULTS_DIR` environment variable to use a
different input or output directory, respectively.

```{r setup, include=FALSE}
required_packages <- c("dplyr", "ggplot2", "knitr", "mgcv")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    "Install the following required package(s): ",
    paste(missing_packages, collapse = ", ")
  )
}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)

data_dir <- Sys.getenv("U1SS_DATA_DIR", unset = "data")
results_dir <- Sys.getenv("U1SS_RESULTS_DIR", unset = "results")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

input_files <- c(
  nearest_u1_sites = file.path(data_dir, "nearest_10_U1_sites_by_gene.RData"),
  u1_summary = file.path(data_dir, "down1000_sense_score0 or more.txt"),
  tissue_expression = file.path(data_dir, "rna_tissue_consensus.tsv")
)

missing_files <- input_files[!file.exists(input_files)]
if (length(missing_files) > 0L) {
  stop(
    "The following input file(s) were not found: ",
    paste(missing_files, collapse = ", ")
  )
}

selected_tissues <- c(
  "cerebral cortex",
  "heart muscle",
  "liver",
  "kidney",
  "lung",
  "spleen",
  "skin",
  "adipose tissue",
  "ovary",
  "small intestine"
)

number_nearest_sites <- 5L
expression_upper_quantile <- 0.95
training_sample_size <- 10000L
training_seed <- 321L
number_prediction_bins <- 10L
```

# Prepare U1SS features

For each gene, the mean score and distance are calculated from the five nearest
U1 sites. The number of downstream sense-oriented U1 sites within 1 kb is taken
from the U1SS summary table.

```{r prepare-u1-features}
u1_environment <- new.env(parent = baseenv())
loaded_objects <- load(input_files[["nearest_u1_sites"]], envir = u1_environment)

if (!"nearest_U1sitelist" %in% loaded_objects) {
  stop("The nearest-site file must contain an object named `nearest_U1sitelist`.")
}

nearest_u1_site_list <- get("nearest_U1sitelist", envir = u1_environment)

required_nearest_columns <- c("gene", "u1_score", "distance")
invalid_entries <- which(!vapply(
  nearest_u1_site_list,
  function(x) {
    is.data.frame(x) &&
      all(required_nearest_columns %in% names(x)) &&
      nrow(x) >= number_nearest_sites
  },
  logical(1)
))

if (length(invalid_entries) > 0L) {
  stop(
    "Every element of `nearest_U1sitelist` must contain at least ",
    number_nearest_sites,
    " rows and the columns: ",
    paste(required_nearest_columns, collapse = ", "),
    ". Invalid element(s): ",
    paste(utils::head(invalid_entries, 10L), collapse = ", ")
  )
}

summarize_nearest_sites <- function(site_table, n_sites) {
  nearest_sites <- utils::head(site_table, n_sites)

  data.frame(
    gene = as.character(nearest_sites$gene[[1L]]),
    mean_score = mean(as.numeric(nearest_sites$u1_score), na.rm = TRUE),
    mean_dist = mean(as.numeric(nearest_sites$distance), na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

nearest_u1_summary <- dplyr::bind_rows(lapply(
  nearest_u1_site_list,
  summarize_nearest_sites,
  n_sites = number_nearest_sites
)) |>
  dplyr::filter(nzchar(.data$gene))

if (any(!is.finite(nearest_u1_summary$mean_score)) ||
    any(!is.finite(nearest_u1_summary$mean_dist))) {
  stop("Non-finite mean U1SS scores or distances were found.")
}

if (anyDuplicated(nearest_u1_summary$gene)) {
  stop("Nearest-site data contain duplicated non-empty gene symbols.")
}

u1_summary_raw <- utils::read.delim(
  input_files[["u1_summary"]],
  check.names = FALSE,
  stringsAsFactors = FALSE
)

required_u1_columns <- c("gene", "name", "num_u1")
if (!all(required_u1_columns %in% names(u1_summary_raw))) {
  stop(
    "The U1SS summary file must contain the columns: ",
    paste(required_u1_columns, collapse = ", ")
  )
}

u1_counts <- u1_summary_raw |>
  dplyr::transmute(
    ens = as.character(.data$gene),
    gene = as.character(.data$name),
    num_u1 = as.numeric(.data$num_u1)
  ) |>
  dplyr::filter(nzchar(.data$gene))

if (any(!is.finite(u1_counts$num_u1))) {
  stop("The `num_u1` column contains missing or non-numeric values.")
}

if (anyDuplicated(u1_counts$gene)) {
  stop("The U1SS summary contains duplicated non-empty gene symbols.")
}

u1_features <- nearest_u1_summary |>
  dplyr::inner_join(u1_counts, by = "gene") |>
  dplyr::select("ens", "gene", "num_u1", "mean_score", "mean_dist")

if (anyDuplicated(u1_features$ens)) {
  stop("U1SS feature data contain duplicated Ensembl gene identifiers.")
}

knitr::kable(utils::head(u1_features), digits = 3)
```

# Prepare mean tissue expression

Normalized transcript abundance (`nTPM`) is averaged across the ten tissues
listed in the setup section. Genes without a matching expression measurement
are assigned an `nTPM` of zero, consistent with the original analysis.

```{r prepare-expression}
tissue_expression_raw <- utils::read.delim(
  input_files[["tissue_expression"]],
  check.names = FALSE,
  stringsAsFactors = FALSE
)

required_expression_columns <- c("Gene", "Gene name", "Tissue", "nTPM")
if (!all(required_expression_columns %in% names(tissue_expression_raw))) {
  stop(
    "The tissue-expression file must contain the columns: ",
    paste(required_expression_columns, collapse = ", ")
  )
}

mean_tissue_expression <- tissue_expression_raw |>
  dplyr::filter(.data$Tissue %in% selected_tissues) |>
  dplyr::transmute(
    ens = as.character(.data$Gene),
    gene = as.character(.data[["Gene name"]]),
    nTPM = as.numeric(.data$nTPM)
  ) |>
  dplyr::group_by(.data$ens, .data$gene) |>
  dplyr::summarise(nTPM = mean(.data$nTPM, na.rm = TRUE), .groups = "drop")

observed_tissues <- unique(tissue_expression_raw$Tissue)
missing_tissues <- setdiff(selected_tissues, observed_tissues)
if (length(missing_tissues) > 0L) {
  stop(
    "The following selected tissue(s) are absent from the expression file: ",
    paste(missing_tissues, collapse = ", ")
  )
}

analysis_data <- u1_features |>
  dplyr::left_join(mean_tissue_expression, by = c("ens", "gene")) |>
  dplyr::mutate(nTPM = dplyr::coalesce(.data$nTPM, 0))

if (any(!is.finite(analysis_data$nTPM)) || any(analysis_data$nTPM < 0)) {
  stop("Expression values must be finite and non-negative.")
}

expression_cutoff <- stats::quantile(
  analysis_data$nTPM,
  probs = expression_upper_quantile,
  names = FALSE,
  na.rm = TRUE
)

analysis_data <- analysis_data |>
  dplyr::filter(.data$nTPM <= expression_cutoff)

if (nrow(analysis_data) < training_sample_size) {
  stop(
    "Only ", nrow(analysis_data),
    " genes remain after filtering; at least ", training_sample_size,
    " are required for the training sample."
  )
}

cat("Expression cutoff (95th percentile):", expression_cutoff, "nTPM\n")
cat("Genes retained for analysis:", nrow(analysis_data), "\n")
```

# Fit the generalized additive model

The training set comprises 10,000 genes selected with a fixed random seed. A
negative-binomial GAM with a log link models mean expression using smooth terms
for U1SS count, mean score, and mean distance.

```{r fit-gam}
set.seed(training_seed)
training_rows <- sample.int(nrow(analysis_data), size = training_sample_size)

analysis_data <- analysis_data |>
  dplyr::mutate(
    data_partition = ifelse(
      dplyr::row_number() %in% training_rows,
      "training",
      "held-out"
    )
  )

training_data <- analysis_data[training_rows, , drop = FALSE]

gam_model <- mgcv::gam(
  nTPM ~ s(num_u1) + s(mean_score) + s(mean_dist),
  data = training_data,
  family = mgcv::nb(link = "log")
)

summary(gam_model)
```

```{r gam-smooths}
plot(gam_model, pages = 1, se = TRUE, shade = TRUE)
```

```{r gam-diagnostics}
set.seed(training_seed)
mgcv::gam.check(gam_model)
```

# Predict expression and assign prediction deciles

Predictions are generated for both training and held-out genes. Genes are then
divided into ten equally sized groups according to predicted expression; bin 1
contains the lowest predictions and bin 10 contains the highest.

```{r predict-and-bin}
analysis_data <- analysis_data |>
  dplyr::mutate(
    predicted_nTPM = as.numeric(stats::predict(
      gam_model,
      newdata = analysis_data,
      type = "response"
    )),
    U1SS_bin = dplyr::ntile(.data$predicted_nTPM, number_prediction_bins)
  )

prediction_plot <- ggplot2::ggplot(
  analysis_data,
  ggplot2::aes(
    x = factor(.data$U1SS_bin, levels = seq_len(number_prediction_bins)),
    y = log10(.data$nTPM + 1),
    fill = factor(.data$U1SS_bin)
  )
) +
  ggplot2::geom_boxplot(notch = TRUE, outlier.size = 0.4) +
  ggplot2::labs(
    x = "Predicted expression decile",
    y = expression(log[10](nTPM + 1))
  ) +
  ggplot2::guides(fill = "none") +
  ggplot2::theme_classic(base_size = 12)

prediction_plot
```

# Save reproducibility outputs

```{r save-results}
prediction_table <- analysis_data |>
  dplyr::select(
    "ens",
    "gene",
    "nTPM",
    "num_u1",
    "mean_score",
    "mean_dist",
    "data_partition",
    "predicted_nTPM",
    "U1SS_bin"
  ) |>
  dplyr::arrange(.data$U1SS_bin, .data$predicted_nTPM)

utils::write.table(
  prediction_table,
  file = file.path(results_dir, "gam_gene_predictions.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

ggplot2::ggsave(
  filename = file.path(
    results_dir,
    "gam_observed_expression_by_prediction_decile.pdf"
  ),
  plot = prediction_plot,
  width = 7,
  height = 5,
  units = "in"
)

saveRDS(gam_model, file.path(results_dir, "gam_model.rds"))
writeLines(
  capture.output(summary(gam_model)),
  file.path(results_dir, "gam_model_summary.txt")
)
```

# Session information

```{r session-information}
sessionInfo()
```
