---
title: "Genome-wide distribution of U1 splice-site scores"
output:
  html_document:
    toc: true
    number_sections: true
---

This document quantifies strand-matched U1 splice-site score (U1SS) intervals
across promoter, exon, intron, and distal intergenic regions on the human
autosomes and sex chromosomes. Genes with at least five annotated exons are
included so that exon and intron ranks 1--4 can be compared with later ranks.

Expected input files in `data/`:

- `all_GENCODE_basic.tsv`: transcript and exon coordinates.
- `map_5ss_hg38_single_nt_plus_corrected.bw`: plus-strand U1SS scores.
- `map_5ss_hg38_single_nt_minus_corrected.bw`: minus-strand U1SS scores.

Set `U1SS_DATA_DIR` or `U1SS_RESULTS_DIR` to use different input or output
directories.

```{r setup, include=FALSE}
required_packages <- c(
  "dplyr",
  "GenomeInfoDb",
  "GenomicRanges",
  "ggplot2",
  "IRanges",
  "knitr",
  "rtracklayer",
  "S4Vectors"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    "Install the following required package(s): ",
    paste(missing_packages, collapse = ", ")
  )
}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

data_dir <- Sys.getenv("U1SS_DATA_DIR", unset = "data")
results_dir <- Sys.getenv("U1SS_RESULTS_DIR", unset = "results")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

input_files <- c(
  annotation = file.path(data_dir, "all_GENCODE_basic.tsv"),
  plus_bigwig = file.path(
    data_dir,
    "map_5ss_hg38_single_nt_plus_corrected.bw"
  ),
  minus_bigwig = file.path(
    data_dir,
    "map_5ss_hg38_single_nt_minus_corrected.bw"
  )
)

missing_files <- input_files[!file.exists(input_files)]
if (length(missing_files) > 0L) {
  stop(
    "The following input file(s) were not found: ",
    paste(missing_files, collapse = ", ")
  )
}

primary_chromosomes <- c(as.character(seq_len(22L)), "X", "Y")
minimum_exon_count <- 5L
promoter_bin_width <- 1000L
number_promoter_bins <- 5L
```

# Import and validate the U1SS tracks

The two bigWig files are explicitly assigned their corresponding strands.
Mitochondrial and non-primary contigs are excluded from both the numerator and
denominator of all density estimates.

```{r import-u1ss}
u1ss_plus <- rtracklayer::import(
  input_files[["plus_bigwig"]],
  format = "BigWig"
)
u1ss_minus <- rtracklayer::import(
  input_files[["minus_bigwig"]],
  format = "BigWig"
)

plus_lengths <- GenomeInfoDb::seqlengths(u1ss_plus)[primary_chromosomes]
minus_lengths <- GenomeInfoDb::seqlengths(u1ss_minus)[primary_chromosomes]

if (any(is.na(plus_lengths)) || any(is.na(minus_lengths))) {
  stop("Both U1SS tracks must contain all chromosomes 1--22, X, and Y.")
}
if (!identical(as.numeric(plus_lengths), as.numeric(minus_lengths))) {
  stop("The plus- and minus-strand bigWig chromosome lengths do not match.")
}

genome_lengths <- stats::setNames(as.numeric(plus_lengths), primary_chromosomes)
primary_seqinfo <- GenomeInfoDb::Seqinfo(
  seqnames = primary_chromosomes,
  seqlengths = genome_lengths
)

u1ss_plus <- GenomeInfoDb::keepSeqlevels(
  u1ss_plus,
  primary_chromosomes,
  pruning.mode = "coarse"
)
u1ss_minus <- GenomeInfoDb::keepSeqlevels(
  u1ss_minus,
  primary_chromosomes,
  pruning.mode = "coarse"
)
GenomicRanges::strand(u1ss_plus) <- "+"
GenomicRanges::strand(u1ss_minus) <- "-"
u1_sites <- c(u1ss_plus, u1ss_minus)

if (length(u1_sites) == 0L) {
  stop("The U1SS tracks contain no intervals on the primary chromosomes.")
}
if (!"score" %in% names(S4Vectors::mcols(u1_sites))) {
  stop("The U1SS bigWig intervals must contain a `score` column.")
}
if (any(!is.finite(as.numeric(S4Vectors::mcols(u1_sites)$score)))) {
  stop("The U1SS tracks contain missing or non-numeric scores.")
}

cat("Strand-specific U1SS intervals:", length(u1_sites), "\n")
```

# Prepare transcript, exon, and intron annotations

Introns are constructed separately within each transcript. This prevents
overlapping transcripts from merging or subtracting one another.

```{r prepare-annotation}
annotation_raw <- utils::read.delim(
  input_files[["annotation"]],
  check.names = FALSE,
  stringsAsFactors = FALSE
)

required_annotation_columns <- c(
  "Gene.ID.version",
  "gene_name",
  "chr",
  "exon_start",
  "exon_end",
  "Exon_rank",
  "transcript_start",
  "transcript_end",
  "strand"
)
if (!all(required_annotation_columns %in% names(annotation_raw))) {
  stop(
    "The annotation file must contain the columns: ",
    paste(required_annotation_columns, collapse = ", ")
  )
}

annotation <- annotation_raw |>
  dplyr::transmute(
    transcript_id = as.character(.data[["Gene.ID.version"]]),
    gene_name = as.character(.data$gene_name),
    chromosome = as.character(.data$chr),
    exon_start = as.integer(.data$exon_start),
    exon_end = as.integer(.data$exon_end),
    exon_rank = as.integer(.data$Exon_rank),
    transcript_start = as.integer(.data$transcript_start),
    transcript_end = as.integer(.data$transcript_end),
    strand = as.character(.data$strand)
  ) |>
  dplyr::filter(
    nzchar(.data$transcript_id),
    .data$chromosome %in% primary_chromosomes
  )

coordinate_columns <- c(
  "exon_start", "exon_end", "exon_rank",
  "transcript_start", "transcript_end"
)
if (any(!stats::complete.cases(annotation[, coordinate_columns]))) {
  stop("The annotation contains missing or non-integer coordinates or ranks.")
}
if (any(!annotation$strand %in% c("+", "-"))) {
  stop("Annotation strand values must be `+` or `-`.")
}
if (any(annotation$exon_start < 1L) ||
    any(annotation$exon_end < annotation$exon_start) ||
    any(annotation$transcript_start < 1L) ||
    any(annotation$transcript_end < annotation$transcript_start)) {
  stop("The annotation contains invalid genomic coordinates.")
}
if (any(annotation$exon_start < annotation$transcript_start) ||
    any(annotation$exon_end > annotation$transcript_end)) {
  stop("One or more exons extend beyond their transcript coordinates.")
}

chromosome_limit <- genome_lengths[annotation$chromosome]
if (any(annotation$exon_end > chromosome_limit) ||
    any(annotation$transcript_end > chromosome_limit)) {
  stop("Annotation coordinates extend beyond the bigWig chromosome lengths.")
}

duplicate_exons <- duplicated(
  annotation[, c("transcript_id", "exon_start", "exon_end")]
)
if (any(duplicate_exons)) {
  stop("The annotation contains duplicated exon coordinates within a transcript.")
}

transcript_consistency <- annotation |>
  dplyr::group_by(.data$transcript_id) |>
  dplyr::summarise(
    chromosomes = dplyr::n_distinct(.data$chromosome),
    strands = dplyr::n_distinct(.data$strand),
    starts = dplyr::n_distinct(.data$transcript_start),
    ends = dplyr::n_distinct(.data$transcript_end),
    .groups = "drop"
  )
if (any(transcript_consistency$chromosomes != 1L) ||
    any(transcript_consistency$strands != 1L) ||
    any(transcript_consistency$starts != 1L) ||
    any(transcript_consistency$ends != 1L)) {
  stop("Each transcript must have one chromosome, strand, start, and end.")
}

annotation <- annotation |>
  dplyr::group_by(.data$transcript_id) |>
  dplyr::filter(dplyr::n() >= minimum_exon_count) |>
  dplyr::ungroup()

if (nrow(annotation) == 0L) {
  stop("No transcripts met the minimum exon-count requirement.")
}

rank_is_valid <- vapply(
  split(annotation$exon_rank, annotation$transcript_id),
  function(ranks) {
    identical(sort(ranks), seq_len(length(ranks)))
  },
  logical(1)
)
if (any(!rank_is_valid)) {
  stop(
    "Exon ranks must be unique and consecutive within each transcript. ",
    "Invalid transcript(s): ",
    paste(utils::head(names(rank_is_valid)[!rank_is_valid], 10L), collapse = ", ")
  )
}

transcript_table <- annotation |>
  dplyr::distinct(
    .data$transcript_id,
    .data$gene_name,
    .data$chromosome,
    .data$transcript_start,
    .data$transcript_end,
    .data$strand
  )

if (anyDuplicated(transcript_table$transcript_id)) {
  stop("Transcript identifiers are not unique after annotation validation.")
}

transcript_ranges <- GenomicRanges::makeGRangesFromDataFrame(
  transcript_table,
  seqnames.field = "chromosome",
  start.field = "transcript_start",
  end.field = "transcript_end",
  strand.field = "strand",
  keep.extra.columns = TRUE
)
GenomeInfoDb::seqinfo(transcript_ranges) <-
  primary_seqinfo[GenomeInfoDb::seqlevels(transcript_ranges)]

make_transcript_introns <- function(exons) {
  exons <- exons[order(exons$exon_start, exons$exon_end), , drop = FALSE]
  intron_start <- utils::head(exons$exon_end, -1L) + 1L
  intron_end <- utils::tail(exons$exon_start, -1L) - 1L

  if (any(intron_end < intron_start)) {
    stop(
      "Overlapping or out-of-order exons were found in transcript: ",
      exons$transcript_id[[1L]]
    )
  }

  number_introns <- length(intron_start)
  intron_rank <- seq_len(number_introns)
  if (exons$strand[[1L]] == "-") {
    intron_rank <- rev(intron_rank)
  }

  data.frame(
    transcript_id = exons$transcript_id[[1L]],
    chromosome = exons$chromosome[[1L]],
    start = intron_start,
    end = intron_end,
    strand = exons$strand[[1L]],
    intron_rank = intron_rank,
    stringsAsFactors = FALSE
  )
}

intron_table <- dplyr::bind_rows(lapply(
  split(annotation, annotation$transcript_id),
  make_transcript_introns
))

make_ranges <- function(data, start_column, end_column) {
  ranges <- GenomicRanges::makeGRangesFromDataFrame(
    data,
    seqnames.field = "chromosome",
    start.field = start_column,
    end.field = end_column,
    strand.field = "strand",
    keep.extra.columns = FALSE
  )
  GenomeInfoDb::seqinfo(ranges) <-
    primary_seqinfo[GenomeInfoDb::seqlevels(ranges)]
  ranges
}

make_exon_rank_ranges <- function(rank_value = NULL, later_than = NULL) {
  selected <- if (!is.null(rank_value)) {
    annotation[annotation$exon_rank == rank_value, , drop = FALSE]
  } else {
    annotation[annotation$exon_rank > later_than, , drop = FALSE]
  }
  make_ranges(selected, "exon_start", "exon_end")
}

make_intron_rank_ranges <- function(rank_value = NULL, later_than = NULL) {
  selected <- if (!is.null(rank_value)) {
    intron_table[intron_table$intron_rank == rank_value, , drop = FALSE]
  } else {
    intron_table[intron_table$intron_rank > later_than, , drop = FALSE]
  }
  make_ranges(selected, "start", "end")
}

cat("Transcripts retained:", nrow(transcript_table), "\n")
cat("Exons retained:", nrow(annotation), "\n")
cat("Introns constructed:", nrow(intron_table), "\n")
```

# Define genomic regions

Promoter bins are non-overlapping 1-kb intervals upstream of each transcript.
Coordinates are clipped to chromosome boundaries. Distal intergenic sequence is
defined separately on each strand after excluding transcripts and their 5-kb
upstream regions.

```{r define-regions}
make_promoter_bin <- function(transcripts, bin_number, bin_width) {
  transcript_strand <- as.character(GenomicRanges::strand(transcripts))
  tss <- ifelse(
    transcript_strand == "+",
    GenomicRanges::start(transcripts),
    GenomicRanges::end(transcripts)
  )

  region_start <- ifelse(
    transcript_strand == "+",
    tss - bin_number * bin_width,
    tss + (bin_number - 1L) * bin_width + 1L
  )
  region_end <- ifelse(
    transcript_strand == "+",
    tss - (bin_number - 1L) * bin_width - 1L,
    tss + bin_number * bin_width
  )

  chromosome <- as.character(GenomicRanges::seqnames(transcripts))
  region_start <- pmax(1, region_start)
  region_end <- pmin(genome_lengths[chromosome], region_end)
  valid <- region_start <= region_end

  ranges <- GenomicRanges::GRanges(
    seqnames = chromosome[valid],
    ranges = IRanges::IRanges(
      start = region_start[valid],
      end = region_end[valid]
    ),
    strand = transcript_strand[valid]
  )
  GenomeInfoDb::seqinfo(ranges) <-
    primary_seqinfo[GenomeInfoDb::seqlevels(ranges)]
  ranges
}

promoter_bins <- lapply(
  seq_len(number_promoter_bins),
  make_promoter_bin,
  transcripts = transcript_ranges,
  bin_width = promoter_bin_width
)

whole_genome <- GenomicRanges::GRanges(
  seqnames = rep(primary_chromosomes, times = 2L),
  ranges = IRanges::IRanges(
    start = 1L,
    end = rep(genome_lengths, times = 2L)
  ),
  strand = rep(c("+", "-"), each = length(primary_chromosomes)),
  seqinfo = primary_seqinfo
)

# `make_promoter_bin()` defines one selected bin. For the complete 5-kb
# promoter, combine the five adjacent 1-kb bins before reducing overlaps.
promoter_5kb <- GenomicRanges::reduce(
  do.call(c, promoter_bins),
  ignore.strand = FALSE
)
genic_and_promoter_ranges <- GenomicRanges::reduce(
  c(transcript_ranges, promoter_5kb),
  ignore.strand = FALSE
)
distal_intergenic <- GenomicRanges::setdiff(
  whole_genome,
  genic_and_promoter_ranges,
  ignore.strand = FALSE
)

region_ranges <- list(
  "Promoter (<=1 kb)" = promoter_bins[[1L]],
  "Promoter (1-2 kb)" = promoter_bins[[2L]],
  "Promoter (2-3 kb)" = promoter_bins[[3L]],
  "Promoter (3-4 kb)" = promoter_bins[[4L]],
  "Promoter (4-5 kb)" = promoter_bins[[5L]],
  "1st Exon" = make_exon_rank_ranges(rank_value = 1L),
  "2nd Exon" = make_exon_rank_ranges(rank_value = 2L),
  "3rd Exon" = make_exon_rank_ranges(rank_value = 3L),
  "4th Exon" = make_exon_rank_ranges(rank_value = 4L),
  "Other Exons" = make_exon_rank_ranges(later_than = 4L),
  "1st Intron" = make_intron_rank_ranges(rank_value = 1L),
  "2nd Intron" = make_intron_rank_ranges(rank_value = 2L),
  "3rd Intron" = make_intron_rank_ranges(rank_value = 3L),
  "4th Intron" = make_intron_rank_ranges(rank_value = 4L),
  "Other Introns" = make_intron_rank_ranges(later_than = 4L),
  "Distal Intergenic" = distal_intergenic
)

empty_regions <- names(region_ranges)[lengths(region_ranges) == 0L]
if (length(empty_regions) > 0L) {
  stop("No intervals were available for: ", paste(empty_regions, collapse = ", "))
}
```

# Calculate U1SS density

Each U1SS interval is counted once for every annotated region it overlaps. The
denominator is the corresponding sum of annotated base pairs, so densities are
annotation-weighted when transcripts overlap.

```{r calculate-density}
summarize_region <- function(region_name, ranges, sites) {
  site_count <- sum(GenomicRanges::countOverlaps(
    ranges,
    sites,
    ignore.strand = FALSE
  ))
  region_size <- sum(IRanges::width(ranges))

  data.frame(
    Region = region_name,
    U1SS_count = site_count,
    Region_size_bp = region_size,
    U1SS_density_per_bp = site_count / region_size,
    stringsAsFactors = FALSE
  )
}

density_table <- dplyr::bind_rows(Map(
  summarize_region,
  names(region_ranges),
  region_ranges,
  MoreArgs = list(sites = u1_sites)
))

region_order <- c(
  "Distal Intergenic",
  "Promoter (4-5 kb)",
  "Promoter (3-4 kb)",
  "Promoter (2-3 kb)",
  "Promoter (1-2 kb)",
  "Promoter (<=1 kb)",
  "1st Exon",
  "1st Intron",
  "2nd Exon",
  "2nd Intron",
  "3rd Exon",
  "3rd Intron",
  "4th Exon",
  "4th Intron",
  "Other Exons",
  "Other Introns"
)

density_table <- density_table |>
  dplyr::mutate(Region = factor(.data$Region, levels = region_order)) |>
  dplyr::arrange(.data$Region)

knitr::kable(density_table, digits = 6)
```

# Plot and save results

```{r plot-density}
region_colors <- c(
  "gray20",
  "rosybrown4", "rosybrown", "rosybrown3", "rosybrown2", "rosybrown1",
  "skyblue1", "navajowhite1",
  "skyblue2", "navajowhite2",
  "skyblue3", "navajowhite3",
  "skyblue4", "navajowhite4",
  "steelblue4", "lightgoldenrod4"
)

density_plot <- ggplot2::ggplot(
  density_table,
  ggplot2::aes(
    x = .data$Region,
    y = .data$U1SS_density_per_bp,
    fill = .data$Region
  )
) +
  ggplot2::geom_col(width = 0.8) +
  ggplot2::scale_fill_manual(values = region_colors, guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = "U1SS density (sites per bp)"
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )
  )

density_plot
```

```{r save-results}
output_table <- density_table |>
  dplyr::mutate(Region = as.character(.data$Region))

utils::write.table(
  output_table,
  file = file.path(results_dir, "u1ss_density_by_genomic_region.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

ggplot2::ggsave(
  filename = file.path(
    results_dir,
    "u1ss_density_by_genomic_region.pdf"
  ),
  plot = density_plot,
  width = 8,
  height = 5,
  units = "in"
)
```

# Session information

```{r session-information}
sessionInfo()
```
